Species motif participation provides unique information about species risk of extinction

### 2023-05-05 manuscript/reorder_main_text.tex
Updates from Overleaf

    % While a relationship between motif participation and persistence is broadly intuitive, so far there has been little research demonstrating which motifs are associated with greater probabilities of persistence.
    % One potential reason for the difficulty in assessing the relationship between species' motif roles and persistence are the large computational resources required to simulate population dynamics.
    % Our solution is to use Bayesian networks, which are much more computationally efficient and facilitate the analysis of larger networks~\citep{Eklof2013,Haussler2020}.
    % One limitation with Bayesian networks is that they only take into account bottom-up effects (see Methods). 
    % Nevertheless, Bayesian networks do capture the vast majority of secondary extinctions obtained from fully-fledged dynamical models including top-down and bottom-up effects~\citep{Eklof2013}, and as such are a good proxy for analysing species persistence following extinctions of primary producers in complex food webs.
    % %The greater computational efficiency and similar [[results? performance? extinctions?]] of Bayesian networks allow us to simulate the effects of a gradient of disturbance strength and thereby explore whether participation in some motifs is associated with greater probability of extinction under strong or weak disturbances.

    %There are 13 possible three-species motifs (the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}, 
    We define motif participation using three-species motifs, which are the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}. 
    Four of these motifs, namely the three-species chain, apparent competition, direct competition, and omnivory, are especially interesting as they are by far the most common in empirical networks~\citep{Stouffer2007, Borrelli2015a, giling2019plant}.
    These four motifs also have clear ecological interpretations.
    The three-species chain indicates vertical flows of energy and biomass in a network and captures the possibility of indirect interactions across multiple trophic levels (e.g., between a plant and a predator of the plant's herbivore).
    The direct competition motif, where two consumers share the same resource, captures the possibility of indirect interactions between consumers (e.g., by depletion of the shared resource). 
    The apparent competition motif, where a predator consumes two resources, can include indirect effects between prey (e.g., by boosting predator populations) in a top-down context, but in a bottom-up context only includes direct interactions.
    Finally, the omnivory motif, where a predator consumes the same resource as its prey, includes captures the potential for a plant to have both direct and indirect effects on the predator. 
    % Furthermore, these four motifs do not include cycles (where species A and B eat each other or where A eats B eats C eats A).
    % As modelling secondary extinctions using Bayesian networks requires acyclic networks (see Methods), these are the only motifs possible.

@@

        We first tested 1) how consumer species persistence varied with motif participation and the severity of disturbance.
        For context, we also test 2) how a consumer's persistence varies with other measures of network structure and a species' place within it.
        As different measures of network structure are not independent since all describe the same network, we also 3) relate species' motif participation to large-scale network structure and to species in-degree and trophic level.
        All analyses are briefly described below and were performed using the R~\citep{R} package \emph{lmerTest}~\citep{lmerTest}.
        % , \emph{MuMIn}~\citep{MuMIn}, and \emph{vegan}~\citep{vegan}.
        See Appendix \emph{3.1 - 3.5} for details.

@@

        To understand the relationships between motif participation and consumers' probabilities of persistence, it is helpful to understand how persistence varies with other aspects of network structure.
        Consumers with higher STL were less likely to persist (Fig.~\ref{fig:lm_CS}, Table S3, \emph{Appendix S6}).
        The relationship between in-degree and persistence depended on the strength of disturbance: higher in-degree was associated with greater persistence when basal species had low probability of extinction but lower persistence when basal species were highly likely to go extinct.

@@

    Identifying the species that are most and least likely to persist after a disturbance is a high priority for conservation efforts~\citep{Bottrilletal2008}.
    In a large set of simulated food webs, a consumer species' probability of persistence after a disturbance affecting basal resources depends on its frequency of participation in the four possible meso-scale motifs.
    Species participating more frequently in the direct competition motif and less frequently in the omnivory motif are more likely to persist, while the effects of participating in the apparent competition or three-species chain motifs depend on the strength of disturbance.
    Together with relationships between persistence and the overall structure of the network~\citep{Dunne2002, Eklof2006, PascualDunne2006,prill2005dynamic, bascompte2005simple} and simpler measures of species' positions within it~\citep{curtsdotter2011robustness, dunne2009cascading, Eklof2006}, these trends can help us to identify vulnerable species within a network.

    % Species-level results summary
    Some of the relationships between motif participation and probability of persistence can be understood by examining relationships between motif participation and other measures of a species' place in the network.
    For example, species with higher trophic levels tended to participate less often in the direct competition motif.
    A high trophic level is always associated with lower persistence since predators necessarily have lower probabilities of persistence than their prey~\citep{Eklof2013}.
    Accordingly, high participation in the direct competition motif, by indicating a low trophic level, is associated with high probability of persistence.
    Similarly, participation in apparent competition decreases with increasing in-degree, and probability of persistence shows the opposite trends with respect to in-degree and participation in apparent competition.

    The effects of other motifs cannot, however, be attributed so straightforwardly to simpler measures of a species' place in the network.
    Like apparent competition, participation in the three-species chain motif decreases with increasing in-degree, yet persistence varied with the frequency of participation in the three-species chain in the same way as in-degree.
    While participation in the three-species chain motif was positively correlated with trophic level, this cannot explain the relationship between trophic level and probability of persistence as trophic level was negatively correlated with persistence across all levels of disturbance.
    Similarly, higher frequencies of participation in the omnivory motif were associated with lower probabilities of persistence, yet the frequency of the omnivory motif increased with in-degree and decreased with trophic level.
    Both high in-degree and low trophic level are associated with higher probability of persistence, except at high levels of disturbance, which would naively suggest that higher proportions of omnivory should also be associated with greater persistence. 
    Considering a species' in-degree, trophic level, \emph{and} its motif role may therefore provide extra information about its extinction risk.

    % Moving to network context
    Relationships between motif participation and persistence can also be placed within the context of relationships between overall network structure, persistence, and the overall frequency of motifs in the network's motif profile.
    For example, since networks with higher connectance contain on average more links per species, it logically follows that they will tend to include a greater proportion of the omnivory motif in the network's motif profile (Fig. SX) and species will tend to participate in the omnivory motif (Fig. 5C). 
    As both higher connectance (Fig. 4C) and more-frequent participation in the omnivory motif (Fig. 2) are generally associated with decreased probability of persistence, it is possible that the negative effect of connectance may be because of the increased frequency of omnivory.

    However, as in other studies, the effect of omnivory on persistence in our networks was highly context-dependent~\citep{bascompte2005simple, Monteiro2016,McLeod2021}.
    Under weak disturbance scenarios, the effect of increased participation in the omnivory motif on consumer persistence within a single network was likely to be neutral or positive (Fig. 3). 
    At high levels of disturbance, however, participating in the omnivory motif was more likely to reduce the probability of persistence.
    These trends also varied with connectance, with participation in the omnivory motif being associated with lower probability of persistence in less-connected networks and higher probability of persistence in more-connected networks as in~\citet{McLeod2021}.
    This is also consistent with earlier work showing greater food-web persistence in networks with higher counts of the omnivory motif~\citep{Stouffer2010b}.
    Highly-connected networks contain many pathways by which a bottom-up disturbance can propagate through the network, leading to stronger effects of disturbance on consumers (Table S6; \emph{Appendix S10}).
    In such a network, it may be especially beneficial for a species to feed on multiple trophic levels and, in a sense, spread its risk of losing prey to bottom-up effects.
    Although species in our model cannot swap prey, the reduced extinction risk for omnivorous species resembles empirical studies showing that adaptive omnivory can increase the permanence of three-species chains~\citep{Fagan1997, Kvrivan2005, AbramsFung2010}.

    [[Still seems relevant but don't know where to fit this in]]
    The stability of omnivory motifs in isolation depends on the strength of the omnivore-resource interaction~\citep{McLeod2021}, but at the whole-network level interactions with other species can stabilize even intrinsically unstable omnivory \citep{Kratina2012}.

    % Caveats - expand on top-down effects a bit to make R2 happy
    There are limitations to the approach presented here.
    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
    Here our focus was to analyze the importance of disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
    For example, one recent study using a dynamical model shows that time to extinction for species following disturbance is tightly coupled to the motifs a species participates in, and that adding top-down effects can lead omnivory to be negatively correlated with persistence \citep{Cirtwill2021_inprep}. 
    If top-down effects are not crucial, the Bayesian network approach could be extended by explicitly incorporating information on species traits.
    In empirical webs, species traits are related to both motif participation~\citep{cirtwill2018feeding} and extinction risk~\citep{Brose2017, curtsdotter2011robustness, Cardillo2005, Purvis2000},
    adding an extra dimension of information which we do not consider here.

    In summary, our results show that participating in different three-species motifs is related to a species' risk of extinction following a bottom-up disturbance, and that these relationships synthesize information provided by other, simpler measures of species roles.
    Critically, our results also underscore the importance of considering the level of disturbance when identifying which species are most at risk of extinction.
    There were strong interactions between the level of disturbance and the relationships between persistence and multiple measures of species roles (motifs and in-degree).
    These simulated results echo other warnings about `tipping points' where previously beneficial traits can become risky~\citep{Latty2019,Golubski2016,Tylianakis2014}.
    For example, at low levels of disturbance a high degree may be protective since species can compensate if one or a few prey are lost. 
    At high disturbance, however, a high-degree species is likely to be affected by disturbances through many paths and have lower persistence.
    These disturbance effects mean that efforts to define a set of target species for conservation interactions must take the current (and future) level of disturbance the community experiences into account.

@@

        \begin{figure}[hb!]
        \centering
        \includegraphics[width=.95\textwidth]{figures/concept_fig_ver2_edit.eps}
        \caption{Conceptual figure of a toy food web. \textbf{a)} This food web contains examples of the four motifs: apparent competition (blue), three-species chain (purple), omnivory (green), and direct competition (yellow). These motifs capture different ways in which disturbances to basal resources can propagate through the network. For example, a disturbance to the bottom species in a three-species chain will have a direct effect on the middle species but only indirectly affect the top species while a disturbance to the bottom species in an omnivory motif will have a direct effect on the middle species and direct and indirect effects on the top species. 
        Note that the nodes participate in more motifs than the ones highlighted and that a species' motif participation vector is the count of all motifs of each type in which the species appears. \textbf{b)} In the same food web, species with identical degree and trophic level can have different motif participation; here species A (light grey node with links in the same motif colored light grey) participates in only two chains while species B (dark grey node with its direct consumption links colored dark grey) participates in chains more than three times more frequently (see inserted table for counts of participation in all four motifs, giving the motif participation vector in parentheses). 
        Despite having equal degree and trophic level, species A may be at greater risk of going extinct than species B because of its dependence on specialist prey species. Two-thirds of the prey of species B are generalists and therefore more likely to survive a disturbance to basal resources. By including information about indirect interactions, motif profiles reflect this difference between species A and B while trophic level and degree do not.}
    \label{fig:concept}
    \end{figure}

>>>>

    %There are 13 possible three-species motifs (the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}, 
    We define motif participation using three-species motifs, which are the most commonly-used size for food web analyses)~\citep{Milo2004,Stouffer2007,Stouffer2012,Cirtwill2015a}. 
    Four of these motifs, namely the three-species chain, apparent competition, direct competition, and omnivory, are especially interesting as they are by far the most common in empirical networks~\citep{Stouffer2007, Borrelli2015a, giling2019plant}.
    These four motifs also have clear ecological interpretations.
    The three-species chain indicates vertical flows of energy and biomass in a network and captures the possibility of indirect interactions across multiple trophic levels (e.g., between a plant and a predator of the plant's herbivore).
    The direct competition motif, where two consumers share the same resource, captures the possibility of indirect interactions between consumers (e.g., by depletion of the shared resource). 
    The apparent competition motif, where a predator consumes two resources, can include indirect effects between prey (e.g., by boosting predator populations) in a top-down context, but in a bottom-up context only includes direct interactions.
    Finally, the omnivory motif, where a predator consumes the same resource as its prey, includes captures the potential for a plant to have both direct and indirect effects on the predator.

@@

        We first tested 1) how consumer species persistence varied with motif participation and the severity of disturbance.
        For context, we also test 2) how a consumer's persistence varies with other measures of network structure and a species' place within it.
        As different measures of network structure are not independent since all describe the same network, we also 3) relate species' motif participation to large-scale network structure and to species in-degree and trophic level.
        All analyses are briefly described below and were performed using the R~\citep{R} package \emph{lmerTest}~\citep{lmerTest}.
        % , \emph{MuMIn}~\citep{MuMIn}, and \emph{vegan}~\citep{vegan}.
        See Appendix \emph{S3} for details.

@@

        To understand the relationships between motif participation and consumers' probabilities of persistence, it is helpful to understand how persistence varies with other aspects of both consumer properties and network structure.

        Consumers with higher trophic level (STL) were less likely to persist (Fig.~\ref{fig:lm_CS}, Table S3, \emph{Appendix S6}).
        The relationship between in-degree and persistence depended on the strength of disturbance: higher in-degree was associated with greater persistence when basal species had low probability of extinction but lower persistence when basal species were highly likely to go extinct.

@@

    Identifying the species that are most and least likely to persist after a disturbance is a high priority for conservation efforts~\citep{Bottrilletal2008}. 
    Network structural properties such as connectance and measures of a species' position in the network such as degree and trophic level are known to affect species' vulnerability~\citep{Dunne2002, Eklof2006, PascualDunne2006,prill2005dynamic, bascompte2005simple,curtsdotter2011robustness, dunne2009cascading, Eklof2006}, but are relatively coarse metrics.
    Defining a species' network role based on its participation in meso-scale motifs provides more nuanced information and may also predict extinction risk~\citep{}.
    Here we tested how the persistence of a consumer in a food web depends on its participation in four different motifs with different levels of disturbance to basal resources.  
    We found that motif participation was related to a consumer's probability of persistence in ways that can't be fully explained by simpler measures of network structure. 
    Species participating more frequently in the direct competition motif and less frequently in the omnivory motif are more likely to persist, while the effects of participating in the apparent competition or three-species chain motifs depend on the strength of disturbance.

    Disturbance strength is often neglected when predicting risk of extinction, but proved to be extremely important in our results.
    Several of the relationships we demonstrated between probability of persistence and species roles changed direction when moving from weak to strong disturbance.
    These simulated results echo other warnings about `tipping points' where previously beneficial traits can become risky~\citep{Latty2019,Golubski2016,Tylianakis2014}.
    For example, at low levels of disturbance a high degree may be protective since species can compensate if one or a few prey are lost. 
    At high disturbance, however, a high-degree species is likely to be affected by disturbances through many paths and have lower persistence.
    This context-dependence contrasts with the general assumption that species with more resources will be less vulnerable~\citep{}.
    These disturbance effects mean that efforts to define a set of target species for conservation interactions must take the current (and future) level of disturbance the community experiences into account.

    % Apparent competition as a case study that motifs are useful
    The relationship between probability of persistence and participation in the apparent competition motif provides a good example of how motifs provide complementary information to that given by simpler measures of a species' network role.
    Species with higher trophic levels tended to participate more frequently in the apparent competition motif.
    As species with higher trophic levels generally have increased risk of going extinct~\citep{Eklof2013,others}, we would therefore expect high frequencies of participation in the apparent competition motif to also be associated with lower probability of persistence.
    However, this is only true for low disturbance levels. 
    For high disturbance, we see the opposite relationship – higher probability of persistence with more-frequent participation in apparent competition.
    This change in the persistence-apparent competition relationship with changing disturbance is what we might expect based on the negative relationship between frequency of participating in apparent competition and in-degree (which showed the opposite trends with respect to persistence).
    However, it is not obvious when trophic level should be the most important predictor of extinction risk and when in-degree should be most important.
    Simpler measures of species' positions in a network could therefore lead us astray either because they provide conflicting predictions or because our expectations are only reasonable for specific types of disturbance.

    % Try to condense the next few bits into one section on omnivory
    Frequency of participation in the omnivory motif likewise provides information about extinction risk that is not easily obtained from in-degree and trophic level.
    Higher frequencies of participation in the omnivory motif were associated with lower probabilities of persistence, yet the frequency of the omnivory motif increased with in-degree and decreased with trophic level.
    Both high in-degree and low trophic level are associated with higher probability of persistence, except at high levels of disturbance, which would naively suggest that higher proportions of omnivory should also be associated with greater persistence. 
    Considering a species' in-degree, trophic level, \emph{and} its motif role may therefore provide extra information about its extinction risk.

    While greater frequency of participation in the omnivory motif was generally associated with reduced probability of persistence across all levels of disturbance, this trend varied strongly with overall network structure (especially connectance).
    Under weak disturbance scenarios, the effect of increased participation in the omnivory motif on consumer persistence within a single network was likely to be neutral or positive (Fig. 3). 
    At high levels of disturbance, however, participating in the omnivory motif was more likely to reduce the probability of persistence as in~\citet{McLeod2021}.
    This is consistent with other studies which have found the effect of omnivory to be highly context-dependent~\citep{bascompte2005simple, Monteiro2016}.

    Highly-connected networks contain many pathways by which a bottom-up disturbance can propagate through the network, leading to stronger effects of disturbance on consumers (Table S6; \emph{Appendix S10}).
    In such a network, it may be especially beneficial for a species to feed on multiple trophic levels and, in a sense, spread its risk of losing prey to bottom-up effects.
    Although species in our model cannot swap prey, the reduced extinction risk for omnivorous species resembles empirical studies showing that adaptive omnivory can increase the permanence of three-species chains~\citep{Fagan1997, Kvrivan2005, AbramsFung2010}.

    In our simulated networks, increased persistence with greater participation in the omnivory motif could have emerged from the requirement that all networks permit the persistence of all species before disturbance was applied.
    In highly-connected networks, which will tend to include more omnivory motifs simply because this motif contains one more link than the others (higher connectance implies more links per species), the abundant omnivory motifs must be generally more stable (or stabilizing) than is necessary for rarer omnivory motifs in less-connected networks.
    Although testing the stability of individual motifs is outside the scope of this study, previous work on suggests that omnivory tends to be more stabilising/stable in isolation when the interactions involved are weak~\citep{}.
    Although omnivory motifs tend to be less stable in isolation the the other motifs we consider here~\citep{Borrelli2015a}, when embedded in a network interactions with other species can stabilise even intrinsically unstable omnivory~\citep{Kratina2012}.

    % Caveats - expand on top-down effects a bit to make R2 happy
    There are limitations to the approach presented here.
    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach have been shown to capture a majority of the secondary extinctions obtained in a fully-fledged dynamical model but doing so with much greater efficiency~\citep{Eklof2013}.
    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2021_inprep}.
    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.
    For this preliminary study, we have also neglected species traits.
    As traits such as body size are related to both motif participation~\citep{cirtwill2018feeding} and extinction risk~\citep{Brose2017, curtsdotter2011robustness, Cardillo2005, Purvis2000} in empirical webs, this is also a potentially useful addition to future work.

    [[Some kind of bridge between caveats and conclusion]]
    Many ecologists, including ourselves, have argued that simple measures of species' roles within food webs are important for understanding species' extinction risk~\citep{}. 
    However, here we show that relationships between extinction risk and measures of a species' place in its food web are likely more complex, varying with the strength of disturbance and the overall network structure.
    This context-dependence means that even limited changes in network structure or the type of disturbance could change which species are most sensitive to disturbance. 
    This argues for further caution in targeted conservation actions and increased acknowledgment of the importance and sensitivity of network structure in ecosystems.

@@

        \begin{figure}[hb!]
        \centering
        \includegraphics[width=.9\textwidth]{figures/concept_fig_ver2_edit.eps}
        \caption{Conceptual figure of a toy food web. \textbf{a)} This food web contains examples of the four motifs: apparent competition (blue), three-species chain (purple), omnivory (green), and direct competition (yellow). These motifs capture different ways in which disturbances to basal resources can propagate through the network. For example, a disturbance to the bottom species in a three-species chain will have a direct effect on the middle species but only indirectly affect the top species while a disturbance to the bottom species in an omnivory motif will have a direct effect on the middle species and direct and indirect effects on the top species. 
        Note that the nodes participate in more motifs than the ones highlighted and that a species' motif participation vector is the count of all motifs of each type in which the species appears. \textbf{b)} In the same food web, species with identical degree and trophic level can have different motif participation; here species A (light grey node with links in the same motif colored light grey) participates in only two chains while species B (dark grey node with its direct consumption links colored dark grey) participates in chains more than three times more frequently (see inserted table for counts of participation in all four motifs, giving the motif participation vector in parentheses). 
        Despite having equal degree and trophic level, species A may be at greater risk of going extinct than species B because of its dependence on specialist prey species. Two-thirds of the prey of species B are generalists and therefore more likely to survive a disturbance to basal resources. By including information about indirect interactions, motif profiles reflect this difference between species A and B while trophic level and degree do not.}
    \label{fig:concept}
    \end{figure}
